Ultrafast Electron Microscopes: Design Criteria, Electron Sources, and Column Modeling

### 2013-05-10 inc/extension/external_forces/formal.tex
tweaks to chapter 3

While the free-space AG model does not contain any external forces, it does include an internal force: the Coulomb force is included to model the repulsion of the pulse's electrons from each other.
As this force is self-influencing, it is the major source of complexity in the model; indeed without this term, a complete analytic solutions is possible, which is identical to ``Gaussian'' optics~\cite{michalik_analytic_2006}.

>>>>

While the free-space AG model does not contain any external forces, it does include an internal force: the Coulomb force is included to model the repulsion of the pulse's electrons from each other.
As this force is self-influencing, it is the major source of complexity in the model; indeed without this term, a complete analytic solutions is possible, which is identical to so-called ``Gaussian optics''~\cite{michalik_analytic_2006}.

### 2013-05-10 inc/extension/external_forces/practical.tex
fixed some operator/matrix symbology
tweaks to chapter 3

\section{Practical Extension to Include Magnetic Lenses, RF Cavities and DC Accelerators}

@@

Of course, physical electron-optical elements occupy only a finite space in the column, and thus their influence on the pulse should be limited to a region near the element itself.
To localize the mathematical result shown in Section \ref{sec:external_forces}, it is convenient to redefine a particular $M_{\alpha}$ to include a super-Gaussian envelope in $z^{\prime}$, the position of the peak of the pulse in the lab frame;
\begin{equation} \label{eq:reg_of_influence}
  M_{\alpha} \to M_{\alpha} \exp \left [ - \left (  \frac{ z^{\prime} - z_{lens}^{\prime} }{ L_{lens} / 2 } \right )^{ 2 n } \right ] \,\text{,}
\end{equation}
where $z_{lens}^{\prime}$ is the position of the center of the lens, $L_{lens}$ is the axial length of the lens and $n$ is the super-Gaussian order parameter.
The order parameter defines the ``sharpness'' of the edge of the region of influence with integer values ranging from 1 (a simple Gaussian) through $\infty$ (a top-hat).
We connect $z^{\prime}$ to $t$ through $ v_{\smallzero} $; the kinetic equations are simply defined for the position of the peak of the pulse.

@@

For the usual case of a spatially compact electron pulse with respect to the dimension of the RF cavity (i.e., $ \Omega z \ll v_{\smallzero} $ and $ \sqrt{\sigma_{\smallT}} \ll a $), \ref{eq:RF_Efield} can be rewritten as
\begin{equation} \label{eq:RF_Efield_approx}
  E ( r , z , z^{\prime} ) = E_{0} \biggl [ \sin \left ( \frac{ \Omega z^{\prime} }{ v_{\smallzero} } + \phi \right ) - \frac{ \Omega z }{ v_{\smallzero} } \cos \left ( \frac{ \Omega z^{\prime} }{ v_{\smallzero} } + \phi \right ) \biggr ] \text{,}
\end{equation}
which demonstrates that the RF cavity acts both as a pulse acceleration (first term on RHS of \ref{eq:RF_Efield_approx}) and a pulse compression or expansion (second term on RHS of \ref{eq:RF_Efield_approx}) device. %TODO clarify this sentance and the next
The linear term (second) then gives the desired form (see \ref{eq:linear_force}) for the force manipulating the temporal dynamics and characteristics during electron pulse compression;
\begin{equation} \label{eq:RF_Force_z}
  \boldsymbol{F}_{ z , RF }^{linear} = - \left [ \frac{ e \Omega E_{0} }{ v_{\smallzero} } \cos \left ( \frac{ \Omega z^{\prime} }{ v_{\smallzero} } + \phi \right ) \right ] z \hat{\operatorname{ \boldsymbol{z} }} \text{.}
\end{equation}

@@

The two main effects contributing to the transverse lens of a $\text{TM}_{010}$-mode RF cavity, its magnetic field component and electric fringe field aperture effects, may also be included within the linear force approximation.
The azimuthal magnetic component of the resonant mode oscillates 90$^{\circ}$ out of phase with the electric component and with a $\operatorname{ J_{1} }$ Bessel radial form; i.e., $ \operatorname{ J_{1} }( k r / a ) \sim k r / a $ for small $r$.
Symmetry and Gauss' Law dictate that in the vicinity of the axial center of the entrance and exit apertures of the RF cavity there exists a radial component of the oscillating electric field determined by $ E_{r} = -\left ( r / 2 \right ) \left ( \partial E_{z} / \partial z \right )_{r=0} $~\cite{kim_rf_1989}, where the form of the axial electric field for $\text{TM}_{010}$ oscillation (\ref{eq:RF_Efield}) implies that, in the context of this analysis, its axial gradient is solely dependent upon the envelope function for the RF cavity (\ref{eq:reg_of_influence}).
In practice, of course, the exact form of $E_{r}(z)$ is strongly tied to the size and shape of the RF cavity apertures (see for example $1 \frac{1}{2}$-cell RF photo-gun designs~\cite{mcdonald_design_1988}). 
For the case of an evacuated cavity, these two contributions result in a net force,
\begin{equation} \label{eq:RF_Force_r}
  \boldsymbol{F}_{ r , RF } = e E_{0} \left [ 
    \frac{ v_{\smallzero} \Omega }{c^{2}} \cos \left ( \frac{ \Omega z^{\prime} }{ v_{\smallzero} } + \phi \right ) + \frac{2 n}{d} \left ( \frac{z^{\prime} - z^{\prime}_{RF}}{d / 2} \right )^{2 n - 1} \sin \left ( \frac{ \Omega z^{\prime} }{ v_{\smallzero} } + \phi \right )
   \right ] r \hat{\operatorname{ \boldsymbol{r} }} \text{.}
\end{equation}
In obtaining \ref{eq:RF_Force_r}, which is also of the form for use in \ref{eq:linear_force}, we have again used an arbitrary super-Gaussian of order $n$ as a region of influence envelope (see \ref{eq:reg_of_influence}) with $L_{RF} \equiv d$, the RF cavity length.
As will be seen in Section \ref{sec:rf_cav_charge}, under the pulse compression conditions ($\phi = 0$ when the pulse is at $z^{\prime} = 0$), both effects are defocussing for the electron pulse and the magnetic contribution is generally smaller than the transverse lenses associated with the RF cavity apertures when $ d = \pi v_{\smallzero} / \Omega $~\cite{kim_rf_1989}, even though its magnitude is maximized for $\phi = 0$.
Other effects, such as magnetic fringe field aperture effects, could also be included, but are significantly weaker.

>>>>

\section{Extension to Include Magnetic Lenses, RF Cavities, and DC Accelerators}

@@

Of course, physical electron-optical elements occupy only a finite space in the column, and thus their influence on the pulse should be limited to a region near the element itself.
To localize the mathematical result shown in Section \ref{sec:external_forces}, it is convenient to redefine a particular $M_{\alpha}$ to include a super-Gaussian envelope in $z^{\prime}$, the position of the peak of the pulse in the lab frame;
\begin{equation} \label{eq:reg_of_influence}
  M_{\alpha} \to M_{\alpha} \exp \left [ - \left (  \frac{ z^{\prime} - z_{lens}^{\prime} }{ L_{lens} / 2 } \right )^{ 2 n } \right ] \,\text{,}
\end{equation}
where $z_{lens}^{\prime}$ is the position of the center of the lens, $L_{lens}$ is the axial length of the lens and $n$ is the super-Gaussian order parameter.
The order parameter defines the ``sharpness'' of the edge of the region of influence with integer values ranging from 1 (a simple Gaussian) through $\infty$ (a top-hat).
We connect $z^{\prime}$ to $t$ through $ v_{\smallzero} $, the speed of the pulse; the kinematic equations are simply defined for the position of the peak of the pulse.

@@

For the usual case of a spatially compact electron pulse with respect to the dimension of the RF cavity (i.e., $ \Omega z \ll v_{\smallzero} $ and $ \sqrt{\sigma_{\smallT}} \ll a $), \ref{eq:RF_Efield} can be rewritten as
\begin{equation} \label{eq:RF_Efield_approx}
  E ( r , z , z^{\prime} ) = E_{0} \biggl [ \sin \left ( \frac{ \Omega z^{\prime} }{ v_{\smallzero} } + \phi \right ) - \frac{ \Omega z }{ v_{\smallzero} } \cos \left ( \frac{ \Omega z^{\prime} }{ v_{\smallzero} } + \phi \right ) \biggr ] \text{,}
\end{equation}
which demonstrates that the RF cavity acts both as a pulse acceleration (first term on RHS of \ref{eq:RF_Efield_approx}) and a pulse compression or expansion (second term on RHS of \ref{eq:RF_Efield_approx}) device. %TODO clarify this sentance and the next
The linear term (second) then gives the desired form (see \ref{eq:linear_force}) for the force manipulating the temporal dynamics and characteristics during electron pulse compression;
\begin{equation} \label{eq:RF_Force_z}
  F_{ z , RF }^{linear} = - \left [ \frac{ e \Omega E_{0} }{ v_{\smallzero} } \cos \left ( \frac{ \Omega z^{\prime} }{ v_{\smallzero} } + \phi \right ) \right ] z \hat{ z } \text{.}
\end{equation}

@@

The two main effects contributing to the transverse lens of a $\text{TM}_{010}$-mode RF cavity, its magnetic field component and electric fringe field aperture effects, may also be included within the linear force approximation.
The azimuthal magnetic component of the resonant mode oscillates 90$^{\circ}$ out of phase with the electric component and with a $\operatorname{ J_{1} }$ Bessel radial form; i.e., $ \operatorname{ J_{1} }( k r / a ) \sim k r / a $ for small $r$.
Symmetry and Gauss' Law dictate that in the vicinity of the axial center of the entrance and exit apertures of the RF cavity there exists a radial component of the oscillating electric field determined by $ E_{r} = -\left ( r / 2 \right ) \left ( \partial E_{z} / \partial z \right )_{r=0} $~\cite{kim_rf_1989}, where the form of the axial electric field for $\text{TM}_{010}$ oscillation (\ref{eq:RF_Efield}) implies that, in the context of this analysis, its axial gradient is solely dependent upon the envelope function for the RF cavity (\ref{eq:reg_of_influence}).
In practice, of course, the exact form of $E_{r}(z)$ is strongly tied to the size and shape of the RF cavity apertures (see for example $1 \frac{1}{2}$-cell RF photo-gun designs~\cite{mcdonald_design_1988}). 
For the case of an evacuated cavity, these two contributions result in a net force,
\begin{equation} \label{eq:RF_Force_r}
  F_{ r , RF } = e E_{0} \left [ 
    \frac{ v_{\smallzero} \Omega }{c^{2}} \cos \left ( \frac{ \Omega z^{\prime} }{ v_{\smallzero} } + \phi \right ) + \frac{2 n}{d} \left ( \frac{z^{\prime} - z^{\prime}_{RF}}{d / 2} \right )^{2 n - 1} \sin \left ( \frac{ \Omega z^{\prime} }{ v_{\smallzero} } + \phi \right )
   \right ] r \hat{ r } \text{.}
\end{equation}
In obtaining \ref{eq:RF_Force_r}, which is also of the form for use in \ref{eq:linear_force}, we have again used an arbitrary super-Gaussian of order $n$ as a region of influence envelope (see \ref{eq:reg_of_influence}) with $L_{RF} \equiv d$, the RF cavity length.
As will be seen in Section \ref{sec:rf_cav_charge}, under the pulse compression conditions ($\phi = 0$ when the pulse is at $z^{\prime} = 0$), both effects are defocussing for the electron pulse and the magnetic contribution is generally smaller than the transverse lenses associated with the RF cavity apertures when $ d = \pi v_{\smallzero} / \Omega $~\cite{kim_rf_1989}, even though its magnitude is maximized for $\phi = 0$.
Other effects, such as magnetic fringe field aperture effects, could also be included, but are significantly weaker.

### 2013-05-10 inc/extension/implementation/implementation.tex
tweaks to chapter 3

A simulation script is composed of creating instances of the classes that represent the column, the pulse, electron-optical elements like magnetic lenses and RF cavities.
These objects act as containers for the information about the physical object that they represent.
Further they manage programmatic tasks related to those objects.
For example, electron-optical element objects manage the Gaussian region-of-influence (\ref{eq:reg_of_influence}) and calculate the force that their fields apply to the pulse.
These objects are managed by a central ``simulation'' class (called \verb!Physics::UEMColumn!) which handles the time evolution of the system.
This arrangement allows for clean, manageable and extensible code.

@@

Additionally, a development version of the bin model presented in Section \ref{sec:bin_model} is available at \url{https://github.com/jberger/Physics-Photoemission}.
Once I am sufficiently confident in its implementation, and more importantly its veracity, I intend to release it to the CPAN as well, most likely under the name \verb!Physics::Photoemission!.
This module depends on the much larger Perl project PDL (\url{http://pdl.perl.org}), a matrix-oriented numerical library, making it slightly more difficult to install, however the speed and implementational simplicity benefit added by that dependency is marked.
Of particular interest to users of \verb!Physics::UEMColumn!, \verb!Physics::Photoemission! also includes a subclass of \verb!Physics::UEMColumn::Photoemission! which uses this binning model to generate the pulse's $z$ direction initial conditions.

>>>>

A simulation script is composed of creating instances of the classes that represent the column, the pulse, electron-optical elements like magnetic lenses and RF cavities.
These objects act as containers for the information about the physical object that they represent.
Further they manage programmatic tasks related to those objects.
For example, electron-optical element objects manage the Gaussian region-of-influence (\ref{eq:reg_of_influence}) and calculate the force that their fields apply to the pulse.
While the pulse object may be created manually, the toolkit contains the class \verb!Physics::UEMColumn::Photocathode! which implements the simple initial conditions described in Section \ref{sec:initial_conditions}.
These objects are managed by a central ``simulation'' class (called \verb!Physics::UEMColumn!) which handles the time evolution of the system.
This arrangement allows for clean, manageable and extensible code.

@@

Additionally, a development version of the bin model presented in Section \ref{sec:bin_model} is available at \url{https://github.com/jberger/Physics-Photoemission}.
Once I am sufficiently confident in its implementation, and more importantly its veracity, I intend to release it to the CPAN as well, most likely under the name \verb!Physics::Photoemission!.
This module depends on the much larger Perl project PDL (\url{http://pdl.perl.org}), a matrix-oriented numerical library, making it slightly more difficult to install, however the speed and implementational simplicity benefit added by that dependency is marked.
Of particular interest to users of \verb!Physics::UEMColumn!, \verb!Physics::Photoemission! also includes a subclass of \verb!Physics::UEMColumn::Photocathode! which uses this binning model to generate the pulse's $z$ direction initial conditions.

### 2013-05-10 inc/extension/initial_conditions/initial_conditions.tex
tweaks to chapter 3

The initial transverse properties of the emitted electron beam are easily understood.
The electron pulse's initial spatial variance is simply
\begin{equation}
\sigma_{\smallT } ( 0 ) = ( \Delta x )^{2} = ( \Delta y)^{2} = \frac{ w^{2} }{ 2 } \punct{.}
\end{equation}
For spatially-uniform photoemission from a planar photocathode, from symmetry we have $ \gamma_{\smallT} ( 0 ) = 0 $; i.e., there is no initial transverse spatial momentum chirp across the electron bunch.
Finally, thanks to the careful analysis of Dowell and Schmerge~\cite{dowell_quantum_2009} (confirmed by Jensen \textit{et. al.}~\cite{jensen_emittance_2010}), we know that metallic photocathode's initial transverse momentum variance at $ t = 0 $ is given by
\begin{equation} \label{eq:jensen_eta}
  \eta_{ \smallT } ( 0 )  = (\Delta p_{\smallT})^2 \approx \frac{ m }{ 3 } ( \hbar \omega - \Phi ) \punct{,}
\end{equation}
in the limit where the maximum excess photoemission energy, $ \Delta E = \hbar \omega - \Phi $ is significantly greater than the thermal energy $ k_{B} T $ of the electrons in the photocathode -- a condition that is generally met in laser-driven DC photoelectron guns.
For completeness we note that for thermionic emission $\Delta p_{\smallT} = \sqrt{m k_B T_e}$ where $k_B$ is the Boltzmann constant and $T_e$ is the electron temperature~\cite{dowell_quantum_2009,jensen_emittance_2010}..

@@

% dc dominated
In the long-pulse regime, the DC field dominated case, we assume that the intrinsic velocity of the emitted electrons is negligible.
The emitted electrons travel an average distance $ \Delta z $, during the $ \Delta t_{ L } $ rms laser pulse duration given by,
\begin{equation}
  \sqrt{ \sigma_{ z } ( 0 ) } \approx \Delta z = \frac{ 1 }{ 2 } a ( \Delta t_{L} )^{ 2 }  = \frac{ e E_{ DC } \tau^{ 2 } }{ 4 m } \punct{.}
\end{equation}
By a similar logic, we can say that the `front' of the pulse then has a velocity equal to $ a \Delta t_{ L } $, while the `back' of the pulse (a distance $ \frac{ 1 }{ 2 } \sqrt{ \sigma_{ z } ( 0 ) } $ from the electron bunch peak) has zero velocity since it has just been generated. From this one sees that the initial linear chirp across the pulse is given by 
\begin{equation}
  \gamma_{ z } ( 0 ) \approx \Delta v \Delta z = m a \Delta t_{ L } \sqrt{ \sigma_{ z } ( 0 ) } = \frac{ e E_{ DC } \tau }{ \sqrt{ 2 } }\sqrt{ \sigma_{ z } ( 0 ) } \punct{.}
\end{equation}

@@

We recognize that these initial conditions for an electron pulse generated by a laser-driven DC gun are idealized in that they neglect several important effects; in particular, the interplay of dispersive ($ \eta_{i}(0) \neq 0 $) and space-charge effects in the acceleration region of the photoelectron gun and transverse and longitudinal effects due to the divergence of the DC field in the vicinity of the gun anode~\cite{berger_dc_2009,togawa_ceb6_2007}.
Such effects and the limits of the validity of the longitudinal terms of \ref{eq:summary}, within the context of the AG model, will be presented in Section \ref{sec:bin_model}.
However, \ref{eq:summary} also represent a defined set of initial conditions that allows the effect of electron optics (i.e., magnetic lenses and RF cavities) on electron pulse propagation to be more readily understood.
Moreover, they reflect the fact that the majority of DC photoelectron guns are driven by laser pulses with spatial and temporal profiles that are close to Gaussian~\cite{williamson_clocking_1997,sciaini_electronic_2009}.
Only for specialized cases, such as the generation of uniform ellipsoid electron bunch shapes, is spatial and temporal manipulation of the incident laser pulse considered~\cite{luiten_how_2004,li_generating_2008}.

>>>>

The initial transverse properties of the emitted electron beam are easily understood.
The electron pulse's initial spatial variance is simply
\begin{equation}
\sigma_{\smallT } ( 0 ) = ( \Delta x )^{2} = ( \Delta y)^{2} = \frac{ w^{2} }{ 2 } \punct{.}
\end{equation}
For spatially-uniform photoemission from a planar photocathode, from symmetry we have $ \gamma_{\smallT} ( 0 ) = 0 $; i.e., there is no initial transverse spatial momentum chirp across the electron bunch.
Finally, thanks to the careful analysis of Dowell and Schmerge~\cite{dowell_quantum_2009} (confirmed by Jensen \textit{et. al.}~\cite{jensen_emittance_2010}), we know that metallic photocathode's initial transverse momentum variance at $ t = 0 $ is given by
\begin{equation} \label{eq:jensen_eta}
  \eta_{ \smallT } ( 0 )  = (\Delta p_{\smallT})^2 \approx \frac{ m }{ 3 } ( \hbar \omega - \Phi ) \punct{,}
\end{equation}
in the limit where the maximum excess photoemission energy, $ \Delta E = \hbar \omega - \Phi $ is significantly greater than the thermal energy $ k_{B} T $ of the electrons in the photocathode -- a condition that is generally met in laser-driven DC photoelectron guns.
For completeness we note that for thermionic emission $\Delta p_{\smallT} = \sqrt{m k_B T_e}$ where $k_B$ is the Boltzmann constant and $T_e$ is the electron temperature~\cite{dowell_quantum_2009,jensen_emittance_2010}.

@@

% dc dominated
In the long-pulse regime, the DC field dominated case, we assume that the intrinsic velocity of the emitted electrons is negligible.
The emitted electrons travel an average distance $ \Delta z $, during the $ \Delta t_{ L } $ rms laser pulse duration given by,
\begin{equation}
  \sqrt{ \sigma_{ z } ( 0 ) } \approx \Delta z = \frac{ 1 }{ 2 } a ( \Delta t_{L} )^{ 2 }  = \frac{ e E_{ DC } \tau^{ 2 } }{ 4 m } \punct{.}
\end{equation}
By a similar logic, we can say that the `front' of the pulse then has a velocity equal to $ a \Delta t_{ L } $, while the `back' of the pulse (a distance $ \frac{ 1 }{ 2 } \sqrt{ \sigma_{ z } ( 0 ) } $ from the electron bunch peak) has zero velocity since it has just been generated. From this one sees that the initial linear chirp across the pulse is given by 
\begin{equation}
  \gamma_{ z } ( 0 ) \approx m \Delta v \Delta z = m a \Delta t_{ L } \sqrt{ \sigma_{ z } ( 0 ) } = \frac{ e E_{ DC } \tau }{ \sqrt{ 2 } }\sqrt{ \sigma_{ z } ( 0 ) } \punct{.}
\end{equation}

@@

We recognize that these initial conditions for an electron pulse generated by a laser-driven DC gun are idealized in that they neglect several important effects; in particular, the interplay of dispersive ($ \eta_{i}(0) \neq 0 $) and space-charge effects in the acceleration region of the photoelectron gun and transverse and longitudinal effects due to the divergence of the DC field in the vicinity of the gun anode~\cite{berger_dc_2009,togawa_ceb6_2007}.
Such effects and the limits of the validity of the longitudinal terms of \ref{eq:summary}, within the context of the AG model, will be presented in Section \ref{sec:bin_model}.
However, \ref{eq:summary} also represents a defined set of initial conditions that allows the effect of electron optics (i.e., magnetic lenses and RF cavities) on electron pulse propagation to be more readily understood.
Moreover, they reflect the fact that the majority of DC photoelectron guns are driven by laser pulses with spatial and temporal profiles that are close to Gaussian~\cite{williamson_clocking_1997,sciaini_electronic_2009}.
Only for specialized cases, such as the generation of uniform ellipsoid electron bunch shapes, is spatial and temporal manipulation of the incident laser pulse considered~\cite{luiten_how_2004,li_generating_2008}.
